---
title: "RobustnessCheck_inter_annotator_agreement"
format: html
editor: visual
author: "Tiziana Ilie"
---

# Robustness check of Kocher's annotation: Inter-annotator agreement

## Set up

```{r}

library(here) # setting the working directory
library(irr) # to calculate the inter-annotator agreement
library(knitr) # to display a useful table

```

## Loading data

```{r}

kocherssample <- read.csv(here::here("data", "annotations", "sample_kocher_germanL1_original.csv"), stringsAsFactors = TRUE)

mysample <- read.csv(here::here("data", "annotations", "sample_kocher_germanL1_annotation.csv"), stringsAsFactors = TRUE)

```

## Calculating Inter-annotator agreement and Cohen's Kappa

I used Cohen's kappa to calculate inter-annotator agreement and interpreted the results using Landis and Koch's scale (1977: 165).

```{r}

Annotated_columns <- merge( #creating a df with only the manually annotated variables
  kocherssample[, c(
    "ID",
    "Text",
    "ReferentialContinuity",
    "SubjectType",
    "ClauseType",
    "InformationStructure",
    "SubjectPosition",
    "VerbConstruction",
    "Verb")],
  mysample[, c(
    "ID",
    "ReferentialContinuity",
    "SubjectType",
    "ClauseType",
    "InformationStructure",
    "SubjectPosition",
    "VerbConstruction",
    "Verb")], 
  by = "ID",
  suffixes = c("_kocher", "_my"))

```

Extracting the cases in which Kocher's and my annotation disagreed, calculating Cohen's K and the annotators agreement in % for each variable:

### SubjectType

```{r}
SubjectType <- "SubjectType"
Disagreement_SubjectType <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(SubjectType, "_kocher")]] != # paste0 joins strings without separator
    Annotated_columns[[paste0(SubjectType, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(SubjectType, "_kocher"), # using the strings as column names
    paste0(SubjectType, "_my")))

CohensK_SubjectType <- kappa2(
  data.frame(
    Annotated_columns$SubjectType_kocher,
    Annotated_columns$SubjectType_my)) #Cohen's κ = 0.887

mean_SubjectType <- mean(
  Annotated_columns$SubjectType_kocher ==
    Annotated_columns$SubjectType_my,
  na.rm = TRUE) * 100 #93.2%

```

### ReferentialContinuity

```{r}

ReferentialContinuity <- "ReferentialContinuity"
Disagreement_ReferentialContinuity <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(ReferentialContinuity, "_kocher")]] !=
    Annotated_columns[[paste0(ReferentialContinuity, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(ReferentialContinuity, "_kocher"),
    paste0(ReferentialContinuity, "_my")))

CohensK_ReferentialContinuity <- kappa2(
  data.frame(
    Annotated_columns$ReferentialContinuity_kocher,
    Annotated_columns$ReferentialContinuity_my)) #Cohen's κ = 0.831

mean_ReferentialContinuity <- mean(
  Annotated_columns$ReferentialContinuity_kocher ==
    Annotated_columns$ReferentialContinuity_my,
  na.rm = TRUE) * 100 #91.5%

```

### ClauseType

```{r}

ClauseType <- "ClauseType"
Disagreement_ClauseType <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(ClauseType, "_kocher")]] !=
    Annotated_columns[[paste0(ClauseType, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(ClauseType, "_kocher"),
    paste0(ClauseType, "_my")))

CohensK_ClauseType <- kappa2(
  data.frame(
    Annotated_columns$ClauseType_kocher,
    Annotated_columns$ClauseType_my)) #Cohen's κ = 0.945

mean_ClauseType <- mean(
  Annotated_columns$ClauseType_kocher ==
    Annotated_columns$ClauseType_my,
  na.rm = TRUE) * 100 #97.5%

```

### InformationStructure

```{r}
InformationStructure <- "InformationStructure"
Disagreement_InformationStructure <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(InformationStructure, "_kocher")]] !=
    Annotated_columns[[paste0(InformationStructure, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(InformationStructure, "_kocher"),
    paste0(InformationStructure, "_my")))

CohensK_InformationStructure <- kappa2(
  data.frame(
    Annotated_columns$InformationStructure_kocher,
    Annotated_columns$InformationStructure_my)) #Cohen's κ = 0.739

mean_InformationStructure <- mean(
  Annotated_columns$InformationStructure_kocher ==
    Annotated_columns$InformationStructure_my,
  na.rm = TRUE) * 100 #95.8%

```

### SubjectPosition

```{r}

SubjectPosition <- "SubjectPosition"
Disagreement_SubjectPosition <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(SubjectPosition, "_kocher")]] !=
    Annotated_columns[[paste0(SubjectPosition, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(SubjectPosition, "_kocher"),
    paste0(SubjectPosition, "_my")))

CohensK_SubjectPosition <- kappa2(
  data.frame(
    Annotated_columns$SubjectPosition_kocher,
    Annotated_columns$SubjectPosition_my)) #Cohen's κ = 1 > full agreement

mean_SubjectPosition <- mean(
  Annotated_columns$SubjectPosition_kocher ==
    Annotated_columns$SubjectPosition_my,
  na.rm = TRUE) * 100 #100.0%

```

### VerbConstruction

```{r}

VerbConstruction <- "VerbConstruction"
Disagreement_VerbConstruction <- subset(
  Annotated_columns,
  Annotated_columns[[paste0(VerbConstruction, "_kocher")]] !=
    Annotated_columns[[paste0(VerbConstruction, "_my")]],
  select = c(
    "ID",
    "Text",
    paste0(VerbConstruction, "_kocher"),
    paste0(VerbConstruction, "_my")))

CohensK_VerbConstruction <- kappa2(
  data.frame(
    Annotated_columns$VerbConstruction_kocher,
    Annotated_columns$VerbConstruction_my)) #Cohen's κ = 0.803

mean_VerbConstruction <- mean(
  Annotated_columns$VerbConstruction_kocher ==
    Annotated_columns$VerbConstruction_my,
  na.rm = TRUE) * 100 #93.2%

```

### Verb

```{r}

Verb <- "Verb"
#saving "Verb" as character instead of a factor as the number of level (unique verbs) is different in both frames
Annotated_columns$Verb_kocher <- as.character(Annotated_columns$Verb_kocher) #'Verb' contains 60 levels here
Annotated_columns$Verb_my <- as.character(Annotated_columns$Verb_my) 
#'Verb' contains 59 levels here

CohensK_Verb <- kappa2(
  data.frame(
    Annotated_columns$Verb_kocher,
    Annotated_columns$Verb_my)) #Cohen's κ = 0.974

mean_Verb <- mean(
  Annotated_columns$Verb_kocher ==
    Annotated_columns$Verb_my,
  na.rm = TRUE) * 100 #97.5%

```

### Displaying inter-annotator agreement results as a table

```{r}

IAA_table <- data.frame(
  Variable = c(
    "SubjectType",
    "ReferentialContinuity",
    "ClauseType",
    "InformationStructure",
    "SubjectPosition",
    "VerbConstruction",
    "Verb"),
  Agreement = c(
    mean_SubjectType,
    mean_ReferentialContinuity,
    mean_ClauseType,
    mean_InformationStructure,
    mean_SubjectPosition,
    mean_VerbConstruction,
    mean_Verb),
  CohensK = c(
    CohensK_SubjectType$value,
    CohensK_ReferentialContinuity$value,
    CohensK_ClauseType$value,
    CohensK_InformationStructure$value,
    CohensK_SubjectPosition$value,
    CohensK_VerbConstruction$value,
    CohensK_Verb$value
  ))

#rounding values for a better overview
IAA_table$Agreement <- round(IAA_table$Agreement, 1)
IAA_table$CohensK <- round(IAA_table$CohensK, 3)

#saveRDS(IAA_table, "2_Robustness_Check/Calculations/IAA_table.rds") #TI: saving as rds to read in thesis file

kable(IAA_table,
  caption = "Inter-annotator agreement and Cohen's Kappa values")

#saveRDS(CohensK_InformationStructure, "2_Robustness_Check/Calculations/CohensK_InformationStructure.rds") #TI: saving as rds to read in thesis file

#saveRDS(CohensK_SubjectPosition, "2_Robustness_Check/Calculations/CohensK_SubjectPosition.rds") #TI: saving as rds to read in thesis file

```

**Concluding remarks**

The re-annotation of a sample of the data yielded high levels of agreement between the original annotation and my annotation. Cohen's K ranged from `{r} round(CohensK_InformationStructure$value, 3)` to `{r} round(CohensK_SubjectPosition$value, 3)`, indicating substantial to almost perfect agreement according to Landis and Koch (1977). These results suggest that the annotation scheme was applied consistently and that the manually annotated variables can be considered robust. Nevertheless, the comments regarding the data exclusion in the `0_Preprocessing.qmd` file should be noted.
