---
title: "4_Extension2_EnglishL1_mupdarfanalysis"
format: html
editor: visual
author: "Tiziana Ilie"
---

# 1. Extension: Investigating the influence of the speaker's gender and the grammatical gender of the referential subject

## Set up and packages

```{r}

library(partykit)
library(dplyr)
library(stringr)
library(lattice)
library(usethis)
library(devtools)
library(moreparty)
library(here)

```

## Loading data

```{r}

datanative<-read.csv(here::here("data", "annotations", "KoRoLi_data", "kocher_native_data_koroli.csv"), stringsAsFactors = TRUE) #loading native speaker data

datal2english<-read.csv(here::here("data", "annotations", "datal2english.csv"), stringsAsFactors = TRUE) #loading English L2 Spanish learner data

```

## STEP 1: Native L1 data predictive model

### Distributions (Duration)

```{r}

sd(datal2english$DurationTask) #15.91207 
range(datal2english$DurationTask) #1-120 
mean(datal2english$DurationTask) #16.99767 (GermanL1 is 68.27338 in comparison!)

```

### MUPDARF

MUPDARF stands for 'Multifactorial Prediction and Deviation Analysis Using Regression/Random Forests'. This approach attempts to improve upon traditional regression- or tree-based methods by firstly training a model on reference speakers (often native speakers in learner corpus studies or British English speakers in variety studies) and then using this model to predict what the reference speaker would produce in the target speaker's situation (often non-native or indigenized variety speakers). The third step consists of determining whether the target speakers made a canonical choice and exploring that variability with a second regression model or classifier (see Gries, 2020).

```{r}

#Before comparing the models, I rename the 'Sex' column 'SpeakerGender', aligning the plot labels with the variables to be investigated in the next section.

datanative <- datanative |> 
  rename("SpeakerGender" = "Sex") #renaming the speaker's gender column
datal2english <- datal2english |> 
  rename("SpeakerGender" = "Sex")

# Final model with SubjectGenus predictor
rfpnative_final <- partykit::cforest(SubjectType ~ SubjectPosition +
                                       InformationStructure +
                                       ReferentialContinuity +
                                       ClauseType +
                                       SubjectGenus, #added SubjectGenus
                                     weights = ifelse(datanative$SubjectType ==
                                                        "dp", 2,
                                               ifelse(datanative$SubjectType ==
                                                        "pron",4,1) ),
                                     data = datanative)

```

### Checking accuracy

```{r}

confusionmatrixrfpnative<-table(predict(rfpnative_final), datanative$SubjectType)
confusionmatrixrfpnative
saveRDS(confusionmatrixrfpnative, "4_Extension2_EnglishGenus/Calculations/confusionmatrix_native_closerepl2.rds") #TI: saving as rds to read in thesis document
sum(diag(confusionmatrixrfpnative))/sum(confusionmatrixrfpnative) #moderate accuracy of 64%, like 64% in Kocher's paper

```

All models provide a similarly moderate level of accuracy. However, the model that includes the *SubjectGenus* variable is equally accurate, so I will select this as the new final model.

### Figure (Variable importance)

```{r}

#pdf(here::here("4_Extension2_EnglishGenus", "Figures", "varimp_native.pdf"), width=5, height = 5)
#varimp<-varimp.cforest(rfpnative_final) #TI: variable importance
#dotplot(sort(varimp),  xlab="Variable importance", col="gray51") 
#dev.off()

#png(here::here("4_Extension2_EnglishGenus", "Figures", "varimp_native.png"), width=500, height = 500)
#varimp<-varimp.cforest(rfpnative_final) #TI: variable importance
#dotplot(sort(varimp),  xlab="Variable importance", col="gray51") 
#dev.off()

```

### Inspecting predictions of the final model (including [SubjectGenus]{.smallcaps})

```{r}

datanative$Predictions<-predict(rfpnative_final) 
#predicts a value of SubjectType for each observation in datanative

datanative$Correct<-as.character(datanative$Predictions) 
#variable called Correct is created and initially contains the predicted class labels

datanative[datanative$Predictions==datanative$SubjectType,]$Correct<-"CORRECT" 
#code compares the predicted value with the true value. When they match, 'Correct' is replaced by 'CORRECT'.

datanative$Correct<-recode(datanative$Correct, pron="predictpron", dp="predictdp",nullsub="predictnullsub") 
#incorrectly classified observations are renamed to make it explicit what the model actually predicted

datanative$Correct<-as.factor(datanative$Correct)

```

Wrong predictions

```{r}

corrtab<-table(datanative$Correct, datanative$SubjectType)
corrtab #different type of visualisation of the confusionmatrix

#DPs
corrtab[1,1]/sum(corrtab[c(1,2,3,4),1]) # 77, originally 80 correct as also in the paper
corrtab[3,1]/sum(corrtab[c(1,2,3,4),1]) #  7, originally 8 null sub
corrtab[4,1]/sum(corrtab[c(1,2,3,4),1]) # 14, originally 10 pronouns

#nullsubs
corrtab[1,2]/sum(corrtab[c(1,2,3,4),2]) #59, originally 60 as also correct in the paper
corrtab[2,2]/sum(corrtab[c(1,2,3,4),2]) #24, originally 24 Dps
corrtab[4,2]/sum(corrtab[c(1,2,3,4),2]) #16, originally 14 pronouns

#pronouns
corrtab[1,3]/sum(corrtab[c(1,2,3,4),3]) #51, originally 38 correct in the paper !!!
corrtab[2,3]/sum(corrtab[c(1,2,3,4),3]) #26, originally 32 Dps
corrtab[3,3]/sum(corrtab[c(1,2,3,4),3]) #21, originally 28 nullsub

```

For L1 English speaking learners of Spanish: More pronouns were correctly predicted using the new model (with *SubjectGenus*) in the L1 English data compared to the L1 German data.

## STEP 2: Running 'Predictnativerf' on the L2German data

Applying the predictive native speaker model to L2 English data:

```{r}

datal2english.mini1 <- datal2english[1:500,]
datal2english.mini1$Predictnativerf<-predict(rfpnative_final, newdata=datal2english.mini1)

datal2english.mini2 <- datal2english[501:1000,]
datal2english.mini2$Predictnativerf<-predict(rfpnative_final, newdata=datal2english.mini2)

datal2english.mini3 <- datal2english[1001:1285,]
datal2english.mini3$Predictnativerf<-predict(rfpnative_final, newdata=datal2english.mini3)

datal2english_new <- bind_rows(list(datal2english.mini1, datal2english.mini2,datal2english.mini3))

#View(datal2english_new)

rm(datal2english) #TI: remove old datal2english
datal2english <- datal2english_new #TI: renaming df
rm(datal2english_new) #TI: remove datal2english_new
```

## STEP 3: Predictions

This step creates a new categorical variable that indicates whether the random forest prediction matches the observed *SubjectType*, and, if not, which category was predicted instead.

### Add CORRECT + predicted value vector

```{r}

#As I annotated the 'unknown' cases that were not included in the predictive model, I received an error message. The following code deselects the unknown cases in order to run:

levels(datal2english$SubjectType)
levels(datal2english$SubjectGenus)

datal2english <- datal2english |> 
  filter(SubjectType != "unknown",
         SubjectGenus != "unknown")|>
  droplevels()

datal2english$ReferenceLikeCat <- as.character(datal2english$Predictnativerf) 
#'ReferenceLikeCat' is created and initially contains the values of 'Predictnativerf'

datal2english[datal2english$Predictnativerf==datal2english$SubjectType,]$ReferenceLikeCat<-"CORRECT" 
#TI: markes correctly predicted cases

datal2english$ReferenceLikeCat<-recode(datal2english$ReferenceLikeCat, pron="predictpron", dp="predictdp",nullsub="predictnullsub") 
#renaming incorrect cases (see above)

datal2english$ReferenceLikeCat<-as.factor(datal2english$ReferenceLikeCat)

datal2english$SubjectType <- factor(datal2english$SubjectType, 
             levels = c("nullsub", "dp", "pron")) #TI: sets the order of factor levels

datal2english$Predictnativerf <- factor(datal2english$Predictnativerf,                                              levels = c("nullsub", "dp", "pron"))

```

### INSPECTION: Predicted vs observed

```{r}

english_pred_actual <- table(datal2english$Predictnativerf, datal2english$SubjectType) #differences in l1 and L2 English
#saveRDS(english_pred_actual, "4_Extension2_EnglishGenus/Calculations/english_pred_actual.rds") #TI: saving as rds to read in thesis document

table(datal2english$ReferenceLikeCat, datal2english$SubjectType)
# how much overlap

predictions_english <- round(table(datal2english$ReferenceLikeCat)[1] /sum(table(datal2english$ReferenceLikeCat)) * 100, digits = 1) # TI: There is a 67% total correct agreement between the native-speaker model and the L2English data.


#nullsub
table(datal2english$Predictnativerf, datal2english$SubjectType)[1,1]/sum(table(datal2english$Predictnativerf, datal2english$SubjectType)[,1]) 
#73.6%
#Kocher: 75% correct (TI: also in table 8 in the paper)

#dps
table(datal2english$Predictnativerf, datal2english$SubjectType)[2,2]/sum(table(datal2english$Predictnativerf, datal2english$SubjectType)[,2]) 
#72.1%
#Kocher: 77% correct

#pronouns
table(datal2english$Predictnativerf, datal2english$SubjectType)[3,3]/sum(table(datal2english$Predictnativerf, datal2english$SubjectType)[,3]) 
#41%
#Kocher: 34% correct

#For L2 speakers, the use of pronouns differs substantially from the patterns learned from native speakers, resulting in fewer correct predictions. For learners of English as a first language, the number is slightly higher.

#In contrast, null subjects and DPs demonstrate a much stronger alignment with the model of a native speaker.


```

## STEP 4: Adding the reference-likeness variable

### Recode Response variable

```{r}

datal2english$ReferenceLike<-as.factor(temp<-ifelse(datal2english$Predictnativerf==datal2english$SubjectType, "TRUE", "FALSE")) 
#creating binary 'ReferenceLike' variable

#MODEL

rfpl2 <- party::cforest(ReferenceLike ~ ClauseType +
                          InformationStructure +
                          ReferentialContinuity + 
                          SubjectType +
                          SubjectPosition + 
                          Score +
                          SubjectGenus, 
                        data = datal2english)
                        #weights = ifelse(datal2german$ReferenceLikeCat =='CORRECT', 1,ifelse(datal2german$ReferenceLikeCat == "predictdp",5,8))
#choice of weights remains unclear
```

### Figure (Proficiency:Score)

```{r}

varimp_l1English <- party::varimp(rfpl2) # Advice ELF: Presumably we need to specify that the party package should be used for this function (as opposed to partykit).

#pdf(here::here("4_Extension2_EnglishGenus", "Figures", "varimp_EnglishL1.pdf"),width=5, height = 5)
#dotplot(sort(varimp_l1English), col="gray51", xlab="Variable importance")
#dev.off() 

#png(here::here("4_Extension2_EnglishGenus", "Figures", "varimp_EnglishL1.png"),width=500, height = 500)
#dotplot(sort(varimp_l1English), col="gray51", xlab="Variable importance")
#dev.off() 
```

### Checking accuracy

```{r}

confusionmatrix_english_closerepl2<-table(predict(rfpl2),datal2english$ReferenceLike)
confusionmatrix_english_closerepl2 #TI: only 18 observations are classified incorrectly (11 in L1 German ds)
#saveRDS(confusionmatrix_english_closerepl2, "4_Extension2_EnglishGenus/Calculations/confusionmatrix_english_closerepl2.rds") #TI: saving as rds to read in thesis document

sum(diag(confusionmatrix_english_closerepl2))/sum(confusionmatrix_english_closerepl2) #98.6% (in L1German 99.8%)

```

### Plotting the surrogate model

```{r}

surro_l2English<-SurrogateTree(rfpl2, maxdepth=4)
Rsquare_surro_closerepli2 <- surro_l2English$r.squared #R^2 is 0.467 (0.999 in L1German ds, 0.945 in the paper and reproducibility part)
#saveRDS(Rsquare_surro_closerepli2, "4_Extension2_EnglishGenus/Calculations/Rsquare_surro_closerepli2.rds")

#FIGURE
#pdf(here::here("4_Extension2_EnglishGenus", "Figures", "surro_l2English.pdf"), width = 21, height = 10) 
#tree <- surro_l2English$tree
#plot(tree)
#dev.off()

#png(here::here("4_Extension2_EnglishGenus", "Figures", "surro_l2English.png"), width = 2000, height = 1000) 
#tree <- surro_l2English$tree
#plot(tree)
#dev.off()
```

Saving the dataframes with the columns *ReferenceLikeCat* and *ReferenceLike* as .csv tables:

```{r}
#View(datal2english)
write.csv(datal2english, here::here("data", "annotations", "withReferenceLike", "ReferenceLike_L1English.csv"), row.names = FALSE)
```
